---
title: "Extra_R_Guide"
author: "Regina"
date: "5/30/2022"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mosaic) # easier analysis functions
library(readr) # easier data import/export
library(tidyr) # Rearranging data
```

## Entering data into R

To enter a new data set (something you will need to do in this course), there are two things you can do (both of which will be illustrated below):

Option 1: If you only have one variable, you can easily type data directly into R. 

Option 2: If you have one or more variables, or a very large data set, you may find it easier to type the data into an Excel spreadsheet, save that spreadsheet as a “.csv” file, and then import that spreadsheet from your computer into R.

We encourage R users to get used to importing a data set directly into R from Excel, since this is usually the easiest way to go.

### Directly typing data into R

Suppose we asked all students in a class (where there are N = 15 students): “How many states have you visited in your lifetime?” The following data was obtained from the class:

15   10   28   27   27   17   29   23   21   36   16   35   32   27   34

To enter this data into R, you would run the code: 
```{r}
states <- c(15,10,28,27,27,17,29,23,21,36,16,35,32,27,34)
```
Of course, it might help here to give you a brief overview of the software before you get too far into R. 

In R, we create objects in R to store information for later use. The object created in the above example is called `states`. We use the assignment operator, `<-`, as a way of assigning information into the object.  Think of the assignment operator `<-` as pointing to the name of the object that we wish to store information in. (Alternatively, instead of `<-` you can also use the equal sign, `=`.) A good programmer will name the objects in a meaningful way in order for easy recall of what is stored in that object. 

The alternative assignment using the `=` would look like:
```{r}
states_duplicate = c(15,10,28,27,27,17,29,23,21,36,16,35,32,27,34)
```

The `c()` function is a way of putting all of the information together in one `vector` (a sequence of multiple values). It essentially concatenates the arguments of the function (each number in this case is an argument of the function) together to form a vector. Arguments are separated by commas in the parentheses. Lastly, R is case sensitive, which means that the symbols `A` and `a` would not refer to the same thing.
 
If we want to see what is assigned under the object named `states`, we can just type in the object name `states` into a code chunk, click Run, and you will see the data contained in that object in the console (which is shown below after the `[1]`):

```{r}
states
```

## Descriptive Statistics

### Summarizing data with numbers

Let's revisit our previous data sets:
```{r}
travels <- read_csv("StatesCountries.csv")
example <- read_csv("Example-Data.csv")
```

In addition to graphing data, we will often want to summarize our data numerically by computing various measures of center (like the mean and median) and variability (like the standard deviation, range, and interquartile range). This can be done by typing each summary function out on different lines of code.   Below are some examples of what this would look like in R.

```{r}
## Mean of states variable
mean(~ states, data = travels)
```

```{r}
## Median of states variable
median(~ states, data = travels)
```

```{r}
## Standard deviation of states variable
sd(~ states, data = travels)
```

```{r}
## Variance of states variable
var(~ states, data = travels)

```

```{r}
## Range of states variable
range(~ states, data = travels)
```

```{r}
## IQR of states variable
IQR(~ states, data = travels)
```

```{r}
## Minimum of states variable
min(~ states, data = travels)
```

```{r}
## Maximum of states variable
max(~ states, data = travels)
```

**Important note if your variable has missing data**: If your variable contains some blanks (because data is missing), when you try to find summary statistics, R will display `NA` rather than giving you the summary statistic. For instance, in the `example` dataset, there are missing values of height:

```{r}
mean(~ height, data = example)
```

For this case, you will need to use `na.rm = TRUE`
For example: 

```{r}
mean(~ height, data = example, na.rm = TRUE)
```

Now, we can see the average height, excluding any missing values.  We do not recommend *always* specifying `na.rm = TRUE`, because it is important for the analyst to *know* that there are missing values.  It's better to only use that option when we absolutely need it.


There is another summary function that gives you the 5-number summary (minimum, 1st quartile, median, 3rd quartile, maximum).   That command is shown below.

```{r}
## Five-number summary of state variable
fivenum(~ states, data = travels)

```

Finally, you can see a summary of several important statistics using the `favstats()` command: five-number summary, mean, standard deviation, number, and number of missing values.

```{r}
favstats(~ states, data = travels)
```

## Describing two categorical variables.

These examples use the `Whickam` dataset from the `mosaicData` package (which was installed when you installed the `mosaic` package). First, we'll load that package and take a quick look at the data:

```{r}
library(mosaicData)
head(Whickham)
```

The `head()` function just shows the first six rows of the data, so we can get a sense of what it looks like.  You can also double-click on the dataset in RStudio to view the full data.

You can find out more about the data (or about any R data or function) by typing the `?` before the dataset.  The help page will show up in the lower right panel. (Warning: for R functions, the help pages can often be too technical for beginners.)

```{r eval=FALSE}
?Whickham
```

We'll compare `outcome` and `smoker`.  We can create a two-way table of counts using the `tally()` function, and save that table as a new object. 

Remember that we save this table as a new object by assigning a name for the table and using `<-` or `=`

```{r}
two_way <- tally(~ outcome + smoker, data = Whickham)
two_way
```

If we want totals, the argument `margins = TRUE` adds on the `Total` rows.
The code below creates a second table (new name) with the row and column totals in the margins.

```{r}
two_way_totals <- tally(~ outcome + smoker, data = Whickham, margins = TRUE)
two_way_totals
```

It can be useful to see the percentages across smokers -- this will tell us what percentage of non-smokers are dead or alive, and what percentage of smokers are dead or alive.  Above, we used the formula `~ smoker + outcome` because we weren't treating either `smoker` or `outcome` as the outcome variable.  Now, we use the formula `outcome ~ smoker` to clarify that `outcome` is our outcome variable, and add `format = "percent"` to show the percentage.

Because we have specified which variable is which, these percentages will tell us the percent of each outcome (alive or dead) grouped by our predictor variable (whether the person smoked or not). What percent of smokers are alive? In the table above, 443 smokers were listed with the `alive` outcome out of 582 total smokers. 443/582 is represented below as 76.116% in the same top right spot. 

```{r}
smoker_percent <- tally(outcome ~ smoker, data = Whickham, format = "percent")
smoker_percent
```

Conversely, if we want to see the percentages across outcomes -- what percentage of dead people are smokers or non-smokers, and what percentage of alive people are smokers or non-smokers -- we use the formula `smoker ~ outcome`:

```{r}
outcome_percent <- tally(smoker ~ outcome, data = Whickham, format = "percent")
outcome_percent
```

Note that the columns in either case add up to 100%.

## Creating side-by-side and segmented bar charts

To create bar graphs, we can use `bargraph`.  Here, we specify the grouping variable by the `groups = ` command. The default result is a side-by-side bar chart:

```{r}
bargraph(~ smoker, groups = outcome, data = Whickham)
```


If we want to make a segmented bar chart, where the groups are stacked on top of each other, simply specify that `stack = TRUE`:


```{r}
bargraph(~ smoker, groups = outcome, data = Whickham, stack = TRUE)
```

## Paired Data

Note that in some data sets, our paired data might be in two separate columns. Think back to our vocab test data, what if instead of having the data in one column, pre scores and the post scores are in separate columns? `t_test` expects just one column to contain all the quantitative values, so we want to rearrange this data. 
Use the **VocabExtra.csv** file to see `gather` in action. We can use the function `gather`, from the `tidyr` package to "gather" all the quantitative values into one column.  
The function looks like this: `newName <- gather(data = originalData, key = pairingVar, value = responseVar, pairGroup1, pairGroup2)`. We are creating a new data frame named `newName` which uses our `originalData` and creates a variable named `pairingVar` which takes on values of `pairGroup1` and `pairGroup2` (which are the column names in `originalData`) and a variable for the `responseVar`. Not that this is general code which is meant to be edited depending on context.

In this context, this function takes the *data* of `vocabExtra`, creates a "key" column which we'll call `timing` to denote whether this is the pre or post score, and then will gather all the scores as "value"s in the column we'll call `score`.  The last arguments are simply the names of the columns we want to gather (`pre` and `post`).

```{r}
vocabExtra <- read_csv("VocabExtra.csv") #read in data
vocabExtra #check out what it looks like

newVocab <- gather(data = vocabExtra, key = timing, value = score, pre, post)
newVocab #take a look at our new object 
```

You will not always have to use `gather()` as some data sets are already organized in this manner, in which case you can skip this step and move on. 

## Chi Squared
What if, instead of having a neat data file, you were given a contingency table like the one below? Here, again, we have the same sample of males and females, and each male and female is categorized as being either Republican (r) or Democrat (d).  Is there a significant relationship between gender and political party affiliation?

|            |	Males |	Females |	Totals|
|------------|---------|-------|---|
| Democrat|	3 |	4 |	7|
|Republican|	4|	4|	8|
|Totals|	7|	8|	15|

We can enter the above information rather easily right into R in order to perform a chi-square analysis.  
We begin by defining our variables (gender and party).  Note we want to enter the information in a particular way here to keep track: we'll first enter the two values for Democrats and then the two values for Republicans, reading left to right across rows.  `ncol` tells R how many columns there are in the table, `nrow` tells R how many rows, and `byrow = TRUE` tells R that we are entering the data going across each row from top to bottom (which many students find more intuitive).

```{r}
gp_data <- matrix(c(3, 4, 4, 4), ncol = 2, nrow = 2, byrow = TRUE)

gp_data
```

Now, we want to add a little bit of information to make this more clear.  The columns of `gp_data` are males and females, while the rows are Democrats and Republicans; we use the R functions `rownames` and `colnames` to assign these values:

```{r}
rownames(gp_data) <- c("D", "R")
colnames(gp_data) <- c("M", "F")

gp_data
```
We can also use the `xchisq.test()` function on this table that we just created, which you'll see generates the same output:

```{r}
xchisq.test(gp_data, correct = FALSE)
```
In terms of the warning message, you can see that we get a message that the chi-square approximation may be incorrect.  This is very likely because of the small expected values, which we can see in the parenthesized text above.
